---
title: "Clustering Wholesale Products"
author: "Zarina"
date: "16/06/2022"
output: github_document
fig_width: 6 
fig_height: 4
---
---
title: "The Application of Unsupervised Learning Techniques in Analysing Wholesale Data"
author: "Zarina"
date: "19/05/2022"
output: github_document
fig_width: 6 
fig_height: 4
---
# Motivation

* Unsupervised learning techniques are of growing importance in a number of fields because these allow interesting patterns and subgroups to be discovered among many observations measured on a set of variables. 

* For instance, in business and marketing, wholesalers might be interested in finding groups of high yield products that are most profitable or have growth potential based on the frequency of transactions. They might also be interested in identifying groups of products that are most frequently purchased together by customers to customise and optimise their sale strategies – e.g., provide a range of offers, prices, promotions, products distributions, or some combinations of marketing segments to increase the chances of products being purchased. Besides that, they might also want to identify groups of customers that share similar profiles, such as purchasing behaviours and histories, so that they can be preferentially shown the products in which they are particularly likely to be interested. 

* Therefore, the main goal of unsupervised learning is not to predict the response 𝑦 as there is no associated response variable 𝑦 in a dataset. Instead, it aims to look for informative patterns and interesting subgroups to mediate the downstream analysis and applications.

## Aim

This report, subsequently, presents the results of the unsupervised learning techniques used – namely, the Principal Components Analysis (PCA) and several clustering algorithms such as K-means/K-medoids, (agglomerative and divisive) hierarchical, and model- based clusterings – in a real dataset which comprises 440 observations corresponding to annual spending in monetary units (m.u) on six product categories (fresh, milk, grocery, frozen, detergents/paper, and delicatessen products). Therefore, the main goal is to discover interesting patterns and homogenous subgroups among these observations to enable the wholesalers to identify interesting products that might be highly correlated to each other or tend to be purchased together, which may further improve the marketing and sale strategies.

# Loading dataset


```{r}
#1. Loading dataset ----
dat <- read.csv("Wholesale_customers_miss.csv",
                sep = ",", header = TRUE)


names(dat)
head(dat)
str(dat)

# change variable name
#names(dat)[names(dat) == "Detergents_Paper"] <- "Detergents.Paper"

```
# Data explorations

## Box plots of ALL variables & outliers 
```{r}

# using ggplot
library(ggplot2)
library("reshape2")
data_long <- melt(dat)
head(data_long)
ggplot(data_long,
       aes(x = variable, y = value, fill = variable)) + ggtitle("Boxplots of all variables")+
  geom_boxplot() + theme(legend.position = "bottom",
                         plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Variable", y = "Value"
                      ) + scale_fill_discrete(name = "Variable")
```
* The highest annual spending is recorded on delicatessen products among all products. This is followed by fresh, detergents/paper, grocery, milk, and frozen products.

* In terms of the distribution of each variable, the annual spending on most of the fresh, milk, and grocery products are normally distributed. On the other hand, the annual spending on frozen and detergents/paper products is slightly right-skewed, possibly driven by extreme outliers. It is also worth noting that two distinct peaks are observed around 5 (m.u) and 10 (m.u) for annual spending on delicatessen products.

*  In addition, each variable appears to have extreme outliers, with most variables having two outliers.

```{r}
## Outliers in Fresh
out <- boxplot.stats(dat$Fresh)$out
idx <- which(dat$Fresh %in% c(out))
dat[idx, c("Fresh")]

## Outliers in Milk
out <- boxplot.stats(dat$Milk)$out
idx <- which(dat$Milk %in% c(out))
dat[idx, c("Milk")]

## Outliers in Grocery
out <- boxplot.stats(dat$Grocery)$out
idx <- which(dat$Grocery %in% c(out))
dat[idx, c("Grocery")]

## Outliers in Frozen
out <- boxplot.stats(dat$Frozen)$out
idx <- which(dat$Frozen %in% c(out))
dat[idx, c("Frozen")]

## Outliers in Detergents_Paper
out <- boxplot.stats(dat$Detergents_Paper)$out
idx <- which(dat$Detergents_Paper %in% c(out))
dat[idx, c("Detergents_Paper")]

## Outliers in Delicatessen
out <- boxplot.stats(dat$Delicatessen)$out
idx <- which(dat$Delicatessen %in% c(out))
dat[idx, c("Delicatessen")]
```

## Histograms 
```{r}

# using ggplot
ggplot(data_long,
       aes(x = value, fill = variable))+
  geom_histogram(aes(y = ..density..)
                 )+
  geom_density(lwd = 0.5, col = "red", alpha = 0.01)+ 
  facet_wrap(~variable, scales = "free") +
  ggtitle("Density Distributions")+
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", h = 0.5))+
  labs(x = "Value", y = "Density") +
  scale_fill_discrete(name = "Variable")


```

## Correlation plot

A correlation matrix plot is also obtained to explore the relationship between the variables in the data set. 

```{r}
library(corrplot)
library(RColorBrewer)
library(psych)

# remove missing values in order to calculate correlation
data <- na.omit(dat)
round(cor(data), 2)

corrplot(cor(data), 
         tl.cex = 0.75, 
         method = "circle", 
         type = "upper", order = "hclust", 
         col = brewer.pal(n=8, name="RdYlBu"), 
         tl.col = "red", 
         diag=FALSE )

pairs.panels(data,
             method = "pearson",
             hist.col = "steelblue",
             density = TRUE,
             ellipses = TRUE)
```
* There are three pairs of variables that are highly, positively correlated with each other:Frozen & Grocery (0.95), Frozen & Fresh (0.79), and Grocery & Fresh (0.86). 

* In contrast, two pairs of variables have moderate, negative correlations. These include Milk & Grocery (-0.43) and Milk & Frozen (- 0.38). 

* The rest of the variables are either entirely not correlated with each other (such as most of the variables with Delicatessen) or very weakly, negatively correlated. 

* Overall, there is evidence that the variables Fresh, Grocery, and Frozen share incredibly redundant information, which needs to be tackled to reduce the redundancy.

## Parallel Coordinate Plot 
```{r}
library(GGally)

ggparcoord(dat,
           alphaLines = 0.2,
           splineFactor = TRUE,
           scale = "uniminmax") +
  scale_color_manual("green") +
 labs(x = "Variable", y = "Value") + ggtitle("Parallel Coordinates Plot") + theme(plot.title = element_text(h = 0.5, face = "bold"))
```

# Missing values analysis 
The distribution of missing data can also be visualized via a matrix plot. 

* Indeed, a more significant portion of missing data (red) occurs in the variable Milk than in Delicatessen. In particular, the missing data pattern displays quite apparent independence where the missing data in Delicatessen appear to be random across all observations in the data set. 

* In addition, the upper block of the matrix where high missing data are observed in Delicatessen seems to occur at higher levels of Milk, which is consistent with the results obtained from the margin plot previously (horizontal red boxplots).
```{r}
library(mice)
library(VIM)

sum(is.na(dat))
md.pattern(dat,  plot = TRUE)
md.pairs(dat)

scattmatrixMiss(dat)

# cor matrix
shadow <- as.data.frame(abs(is.na(dat)))
miss.shadow <- shadow[, which(unlist(lapply(shadow, sum))!=0)]
round(cor(miss.shadow), 3)

round(cor(dat[!names(dat)%in%c("Milk")],
          miss.shadow, use = "pairwise.complete.obs"), 3)
aggr(dat,
     col = mdc(1:2),
     numbers = TRUE,
     sortVars = TRUE,
     labels = names(dat),
     cex.axis = .7,
     gap = 3,
     ylab = c("Proportion of Missingness", "Missingness Pattern"))

marginplot(dat[, c("Milk", "Delicatessen")],
           col = mdc(1:2),
           cex.numbers = 1.2,
           pch = 19)




names(dat)[names(dat) == "Dtrgnt_Ppr"] <- "Dtrgnt_Pr"
names(dat)[names(dat) == "Delicatessen"] <- "Delicatessn"

par(mfrow = c(1, 1))
matrixplot(dat, labels = TRUE, interactive = F, main = "Matrix Plot", sortby = 5)


which(is.na(dat$Milk))


```
# 3.1 Multiple imputation 

For this data set, the imputation analysis involves using all variables as predictors to impute the missing values in the variables Milk and Delicatessen using three methods: 

* Predictive mean matching (PMM), which is based on selecting a data point from the original, non- missing data that has a predicted value close to the predicted value of the missing sample

* Bayesian linear regression (NORM)

* Linear regression ignoring model error (NORM.NOB). 

The imputation is repeated ten times, producing ten complete/imputed data sets (m = 10). The quality of the imputation is compared between different methods used. Only one of the completed data sets that matches the observed data well is used for the subsequent analysis.
```{r}
# with PMM
mice_dat <- mice(dat, m = 10, seed = 123)
summary(mice_dat)

# with (Bayesian) linear regression
mice_dat2 <- mice(dat, m = 10, seed = 123,
                  method = c("","norm.nob","", "", "","norm"))

summary(mice_dat2)
summary(mice_dat2$imp)


## extract the 2nd imputed data set
completed_dat <- complete(mice_dat, 2)
head(completed_dat)

completed_dat2 <- complete(mice_dat2, 2)
head(completed_dat2)

require(lattice)

#xyplots 
library(gridExtra)

xyplot(mice_dat, Milk ~ Delicatessn|as.factor(.imp), pch = 18, cex = 1, main = "PMM")
xyplot(mice_dat2, Milk ~ Delicatessn|as.factor(.imp), pch = 20,
               cex = 1.4, main = "(Bayesian) Linear Regression")



# density plots

densityplot(mice_dat, main = "PMM")
densityplot(mice_dat2, main = "(Bayesian) Linear Regression")

```
# PCA 

The data set has also presented evidence for multicollinearity between variables. Thus, there is an undesirably high chance of information redundancy. Therefore, PCA is used to reduce the dimensionality of the data set by keeping only a small number of dimensions or principal components that still explains most of the variability in the data. 

PCA then allows better visualisation of data using fewer representative variables that still retain as much of the information as possible, thus minimising the effects of multicollinearity and information redundancy for downstream analysis. Otherwise, 15 scatterplots will be needed to visualise 440 observations, which will certainly not be possible to look at all of them; moreover, most likely none of them will be informative since they each contain only a small fraction of the total information present in the data set. Therefore, PCA is performed on this data set as part of an exploratory data analysis, and the results of PCA are then subsequently used in clustering.

## Using Scaled data
To perform PCA, all variables in the imputed data set are first scaled to have mean zero and standard deviation one. This step is considered necessary because performing PCA on unscaled data would cause a greater weightage or importance to be placed onto variables with higher variances than other variables. 

In addition, even though the variables are measured in the same unit (m.u), some products may be purchased more frequently than others. For instance, this data set highlights higher annual spending on delicatessen products than frozen products. If these are not scaled, high-yield purchases like delicatessen products will have a larger effect on inter-products dissimilarities, hence, on the clusters ultimately obtained, than low purchases like frozen products. 

In contrast, if scaling is first performed on the variables, then each variable will, in effect, be given equal importance in the subsequent clustering performed.

```{r}
library(factoextra)
pr.out2 <- prcomp(completed_dat2,
                 scale = TRUE)
# scree plot
fviz_screeplot(pr.out2, addlabels = TRUE, barfill = c("#225ea8", "#1d91c0","#41b6c4",
                                                     "#7fcdbb", "#c7e9b4", "#edf8b1"), 
               barcolor = c("#0c2c84", "#1d91c0","#41b6c4",
                            "#7fcdbb", "#c7e9b4", "#edf8b1")) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Principal Component", y = "Percentage of Variance Explained") + ggtitle("PCA on Scaled Data")



fviz_pca_var(pr.out2,
                   col.var = "contrib", # Color by contributions to the PC
                   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                   repel = TRUE     # Avoid text overlapping
) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)") + ggtitle("Variables - PCA - Scaled Data")
# contribution of variables on scaled data to compare


```
## using unscaled data to compare

```{r}
pr.out <- prcomp(completed_dat2,
                 scale = FALSE)
pr.out

# scree plot
library(factoextra)
fviz_screeplot(pr.out, addlabels = TRUE, barfill = c("#225ea8", "#1d91c0","#41b6c4",
                                                     "#7fcdbb", "#c7e9b4", "#edf8b1"), 
               barcolor = c("#0c2c84", "#1d91c0","#41b6c4",
                            "#7fcdbb", "#c7e9b4", "#edf8b1")) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Principal Component", y = "Percentage of Variance Explained") + ggtitle("PCA on Unscaled Data")

# cumulative var plot
fviz_eig(pr.out)

# contribution of variables on unscaled data
fviz_pca_var(pr.out,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "First Principal Component (47.1%)", y = "Second Principal Component (38.0%)") + ggtitle("Variables - PCA - Unscaled Data")

```
Indeed, the principal loadings are different when the variables are not scaled to have mean zero and standard deviation one, as shown above. Observe that the first principal component places the largest weight on the variable Delicatessen, whereas the second principal component places most of its weight on the variable Detergents/Paper. 

By far, these two variables have the largest variances among all variables (13.17 and 10.94, respectively). Their variances are about ten times higher than the variances of other variables. For example, the variance for the other four variables is comparably very small: Fresh (0.71), Milk (0.21), Grocery (2.86), Frozen (0.47). Therefore, when unscaled, the loading vector places much less weight on these variables. As predicted, the contributions of Delicatessen and Detergents/Paper variables are incredibly high. When comparing this result with the previous one, it can be seen that scaling does indeed have a substantial effect on the results obtained.

## PCA on imputed data 1

```{r}
## 5.1 Using manual ----
## check to use covariance matrix or correlation matrix (scale the data)

sd <- apply(completed_dat2, 2, sd)
sd

var.means <- apply(completed_dat2, 2, mean)
var.means

## There are on average as twice as many Delicatessen as Grocery, and more than 6 times as many Delicatessen as Frozen

var.variance <-apply(completed_dat2, 2, var)
var.variance

## Not suprisingly, the variablese also have vastly different variances: the Delicatessen by far has the largest mean and variance. If we failed to scale the variables before performing PCA, then most of the principal components observed would be driven by the Delicatessen varuable. Thus, it is important to standardize the variables to have mean zero and standard deviation one before performing PCA.

## the variables should be scaled/correlation matrix used

S <- cor(completed_dat2)
eigdec <- eigen(S)
eigdec

# eigenvalues/ variances
eig <- eigdec$values
eig

# eigenvectors/ PC loadings
pc_loading <- eigdec$vectors
rownames(pc_loading) <- colnames(completed_dat2)
pc_loading

# the proportion of variance explained
eig <- eigdec$values
variance <- eig*100/sum(eig)

# cumulative variances
cumvar <- cumsum(variance)
eig2 <- data.frame(eig = eig,
                   variance = variance,
                   cumvariance = cumvar)
eig2

# scree plot
par(mfrow = c(1, 2))
barplot(eig2[, 2],
        names.arg = 1:nrow(eig2),
        xlab = "Principal Component",
        ylab = "Percentage of Variance Explained",
        col = "steelblue")
lines(x = 1:nrow(eig2),
      eig2[, 2],
      type = "b", 
      col = "red",
      pch = 16)

# cumvar scree plot
barplot(eig2[, 3],
        names.arg = 1:nrow(eig2),
        xlab = "Principal Component",
        ylab = "Cumulative Percentage of Variance Explained",
        col = c("#225ea8", "#1d91c0","#41b6c4",
                "#7fcdbb", "#c7e9b4", "#edf8b1"))

lines(x = 1:nrow(eig2),
      eig2[, 3],
      type = "b", 
      col = "red",
      pch = 16)

pc_score = as.matrix(scale(completed_dat))%*% pc_loading
colnames(pc_score) <- paste0("PC", 1:6)
pc_score[1:4,]


pairs.panels(pc_score,
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
)



# Using Factoextra library 


pr.out <- prcomp(completed_dat2,
                 scale = TRUE)
names(pr.out)
summary(pr.out)

#PC loadings
pr.out$rotation

#PC score
pr.out$x
dim(pr.out$x)

# sd of each PC
pr.out$sdev

# scree plot
fviz_screeplot(pr.out, addlabels = TRUE, barfill = c("#225ea8", "#1d91c0","#41b6c4",
                                                      "#7fcdbb", "#c7e9b4", "#edf8b1"), 
                                                     barcolor = c("#0c2c84", "#1d91c0","#41b6c4",
                                                                  "#7fcdbb", "#c7e9b4", "#edf8b1")) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Principal Component", y = "Percentage of Variance Explained")


var <- get_pca_var(pr.out)
var$cor
var$cos2 # quality of representation
var$contrib # contribution of each var to PCs



# Individuals with a similar profile are grouped together 
fviz_pca_ind(pr.out,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)")

fviz_pca_ind(pr.out,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,
             axes = c(2, 3))+ theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Second Principal Component (17.4%)", y = "Third Principal Component (16.0%)")



# graph of variables: positive correlated variables point to the same side  of the plot; negatively correlated variablse point to the opposite sides of the graph

fviz_pca_var(pr.out,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)")

fviz_pca_var(pr.out,
                   col.var = "contrib",
                   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                   repel = TRUE,
                   axes = c(2, 3))+ theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Second Principal Component (17.4%)", y = "Third Principal Component (16.0%)")



# biplot of individuals and variables


fviz_pca_biplot(pr.out, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "cos2",  
                # Individuals color
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)")

fviz_pca_biplot(pr.out, repel = TRUE,
                axes = c(2, 3),
                col.var = "#2E9FDF", # Variables color
                 # Individuals color
                col.ind = "cos2",  
                # Individuals color
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
) + theme(plot.title = element_text(face = "bold", h = 0.5)) + labs(x = "Second Principal Component (17.4%)", y = "Third Principal Component (16.0%)")


# Proportion of variance explained
summary(pr.out)
plot(pr.out)






# Total cos2 of variables on Dim.1 and Dim.3
fviz_cos2(pr.out, choice = "var", axes = 1:2)

## Note that, a high cos2 indicates a good representation of the variable on the principal component. In this case the variable is positioned close to the circumference of the correlation circle.

## A low cos2 indicates that the variable is not perfectly represented by the PCs. In this case the variable is close to the center of the circle.


# a more elegant biplot


fviz_pca_biplot(pr.out,
                label = "var", axes = c(1, 2),
                col.ind = "cos2",
                col.var = "red", repel = TRUE
) + labs(x = "First Principal Component", 
         y = "Second Principal Component",
         title = "PCA biplot of variables and individuals") +
  theme(plot.title =  element_text(hjust = 0.5,
                                   face = "bold")
        ) + scale_color_gradient2(low = "white",
                                  mid = "blue",
                                  high = "red",
                                  midpoint = 0.6)

# color by cos2 values: quality on the factor map
fviz_pca_var(pr.out, 
             col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE) + # Avoid text overlapping 
  labs(x = "First Principal Component", 
  y = "Second Principal Component",
title = "PCA biplot of variables") +
  theme(plot.title =  element_text(hjust = 0.5,
                                   face = "bold"))
```
# Clustering analysis
```{r}
dat.scaled <- scale(completed_dat)
head(dat.scaled)

dat.scaled2 <- scale(completed_dat2)
head(dat.scaled2)
```

# find optimal number of clusters 
```{r}
fviz_nbclust(dat.scaled,
             kmeans,
             method = "wss") +
  theme(plot.title = element_text(h = 0.5,
                                  face = "bold"))+
  geom_vline(xintercept = 4,
             linetype = "dashed",
             color = "steelblue", size = 0.5)

fviz_nbclust(dat.scaled2,
             kmeans,
             method = "wss") + ggtitle("Elbow Method") +
  theme(plot.title = element_text(h = 0.5, face = "bold", size = 12)) 

fviz_nbclust(dat.scaled2,
                   kmeans,
                   method = "silhouette") + ggtitle("Average Silhoutte Method") +
  theme(plot.title = element_text(h = 0.5, face = "bold", size = 12))



# silhouttee plot for k-means
library(cluster)
set.seed(123)
km.res <- kmeans(dat.scaled2,
                 2,
                 nstart = 25)

sil <- silhouette(km.res$cluster, dist(dat.scaled2))

head(sil[, 1:3], 5)

fviz_silhouette(sil, palette = "jco", ggtheme = theme_classic()) + theme(plot.title = element_text(h = 0.5, face = "bold", size = 12))

# k = 5
set.seed(123)
km.res2 <- kmeans(dat.scaled2,
                 5,
                 nstart = 25)

sil2 <- silhouette(km.res2$cluster, dist(dat.scaled2))

head(sil[, 1:3], 5)

fviz_silhouette(sil2, palette = "jco", ggtheme = theme_classic()) + theme(plot.title = element_text(h = 0.5, face = "bold", size = 12))


# compute k-means with k = 2
set.seed(123)
km.res1 <- kmeans(dat.scaled2, 1, nstart = 25)
km.res <- kmeans(dat.scaled2,
                 2,
                 nstart = 25)
set.seed(123)
km.res2 <- kmeans(dat.scaled2,
                 5,
                 nstart = 25)
names(km.res)
km.res
head(km.res$cluster, 4) # cluster no. of each observation

km.res$size # size of clusters

#cluster means
km.res$centers

## the total variability explained by the 5 clusters is 31.5%

# compute the mean of each variables by clusters using original data
aggregate(completed_dat, by = list(cluster = km.res$cluster), mean)

aggregate(completed_dat2, by = list(cluster = km.res2$cluster), mean)

# Visualize kmeans clustering
fviz_cluster(km.res, dat.scaled2, ellipse.type = "norm", palette = "Set2", ggtheme = theme_minimal()) + ggtitle ("K-means clustering with k = 2") + labs(x = "First Principal Component (48.8%)", y = "Second Principal Component (17.0%)") + theme(plot.title = element_text(face = "bold", h = 0.5))



```
## K-means clustering on the first 3 PCs 
```{r}
pr.out <- prcomp(dat.scaled)


set.seed(123)
kmean.out <- kmeans(pr.out$x[, c(1, 2)],
                    centers = 4, 
                    nstart = 25)

fviz_cluster(kmean.out, pr.out$x[,c(1,2)], 
             ellipse.type = "norm") + ggtitle("Cluster Plot with K = 4") +
  labs(x = "First Principal Component (48.8%)", y = "Second Principal Component (17.0%)") +theme(plot.title = element_text(h = 0.5)) 

set.seed(1234)
kmean.out <- kmeans(pr.out$x[, c(2, 3)],
                    centers = 4, 
                    nstart = 25)

fviz_cluster(kmean.out, pr.out$x[,c(2,3)], 
             ellipse.type = "norm")  + ggtitle("Cluster Plot with K = 4") +labs(x = "Second Principal Component (17.0%)", y = "Third Principal Component (16.7%)") +theme(plot.title = element_text(h = 0.5))



```
## K-medoids using PAM algorithms 
```{r}

# optimal number for k medoids
plt1 <- fviz_nbclust(dat.scaled2,
             pam,
             method = "wss") + ggtitle("Elbow Method") +
  theme(plot.title = element_text(h = 0.5, size = 12, 
                                  face = "bold"))
plt2 <- fviz_nbclust(dat.scaled2,
             pam,
             method = "silhouette") +ggtitle("Average Silhoutte Method") +
  theme(plot.title = element_text(h = 0.5, size = 12,
                                  face = "bold"))

library(cluster)
set.seed(123)
pam.res <- pam(dat.scaled2, 2)
print(pam.res) 

dd <- cbind(completed_dat2, cluster = pam.res$clustering)
head(dd, n =3)
pam.res$medoids
```
### compare silhoutte for k means and k medoids
```{r}
set.seed(123)
pam.res <- pam(dat.scaled2, 2)
pam.sil <- silhouette(pam.res$clustering, dist(dat.scaled2))

fviz_silhouette(pam.sil, palette = "jco", ggtheme = theme_classic()) + theme(plot.title = element_text(h = 0.5, face = "bold", size = 12)) + labs(y = "K-medoids Silhouette")

set.seed(123)
km.res <- kmeans(dat.scaled2,
                 2,
                 nstart = 25)

sil <- silhouette(km.res$cluster, dist(dat.scaled2))


fviz_silhouette(sil, palette = "jco", ggtheme = theme_classic()) + theme(plot.title = element_text(h = 0.5, face = "bold", size = 12)) + labs(y = "K-means Silhouette")
```
## visualizing PAM clusters
```{r}
# k= 2
set.seed(123)
pam.res <- pam(dat.scaled2, 2)
fviz_cluster(pam.res, pr.out$x[, c(1, 2)],
             ellipse.type = "norm") + ggtitle("Cluster plot with k = 2") +

  labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)") +theme(plot.title = element_text(h = 0.5, size = 12)) 

fviz_cluster(pam.res, pr.out$x[,c(2,3)], 
             ellipse.type = "norm")  + ggtitle("Cluster plot with k = 2") +labs(x = "Second Principal Component (17.4%)", y = "Third Principal Component (16.0%)") +theme(plot.title = element_text(h = 0.5))

pam.res$id.med

# k = 5
set.seed(123)
pam.res2 <- pam(dat.scaled2, 5)
pam.res2$medoids
fviz_cluster(pam.res2, pr.out$x[, c(1, 2)],
                   ellipse.type = "norm") + ggtitle("Cluster plot with k = 5") +labs(x = "First Principal Component (48.7%)", y = "Second Principal Component (17.4%)") +theme(plot.title = element_text(h = 0.5, size = 12)) 

fviz_cluster(pam.res2, pr.out$x[,c(2,3)], 
                   ellipse.type = "norm")  + ggtitle("Cluster plot with k = 5") +labs(x = "Second Principal Component (17.4%)", y = "Third Principal Component (16.0%)") +theme(plot.title = element_text(h = 0.5))


## rows 146 and 351 are used as the medoids; the columns are the variables

```


# Hierarchical Clustering 

## Optimal number of clusters 
```{r}
#wss method
fviz_nbclust(dat.scaled2, FUN = hcut, method = "wss") + ggtitle("Elbow Method")+
  theme(plot.title = element_text(h = 0.5, face = "bold", size = 12))

# avg silhoutte method
fviz_nbclust(dat.scaled2, FUN = hcut, method = "silhouette") + ggtitle("Average Silhouette Method") +
  theme(plot.title = element_text(h = 0.5, size = 12, face = "bold"))
```
# Agglomerative Clusterings 
```{r}
# load required libraries
library(tidyverse) # data manipulation
library(cluster)  # clustering algorithms
library(factoextra) # clustering visualization
library(dendextend) # for comparing two dendrograms
library(igraph)
# Compute distances
dd <- dist(dat.scaled2, method = "euclidean")

# Complete linkage
hc.complete <- hclust(dd, method = "complete")
hc.complete

# Average linkage
hc.average <- hclust(dd, method = "average")
hc.average

# Single linkage
hc.single <- hclust(dd, method = "single")
hc.single

# Ward Method
hc.ward <- hclust(dd, method = "ward.D2")
hc.ward

# Plot dendrograms for each
par(mfrow = c(2, 2))
plot(hc.complete, main = "Complete Linkage", xlab = "", sub = "", cex = .6, hang = -1)

plot(hc.average, main = "Average Linkage", xlab = "", sub = "", cex = .6, hang = -1)

plot(hc.single, main = "Single Linkage", xlab = "", sub = "", cex = .6, hang = -1)

plot(hc.ward, main ="Ward Method", xlab = "", sub = "", cex = .6, hang = -1)
# Complete Linkage only
par(mfrow = c(1, 1))

xsc <- scale(completed_dat2)

plot(hclust(dist(xsc), method = "complete"), main = "Hierarchical Clustering with Scaled Features", xlab = "", sub = "", cex = .6, hang = -1)

# cut dendrogram at height = 5
hc.clusters <- cutree(hc.complete, 5)
abline(h = 5, col = "red", lty = 3,lwd = 3)

# compute with agnes
hc2 <- agnes(dat.scaled2, method = "complete")

# agglomerative coefficient
hc2$ac # 0.9115044

# methods to assess
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(dat.scaled2, method = x)$ac
}

map_dbl(m, ac)
#average    single  complete      ward 
#0.8205011 0.6646577 0.9115044 0.9744263 

# visualize dendrogram for ward method
par(mfrow = c(1, 1))
hc3 <- agnes(dat.scaled2, method = "ward")
pltree(hc3, cex = 0.6, hang = -1, main = "Dendrogram of Agnes", xlab = "Ward Method", sub = "")
```

## compare cut from kmeans and hc clusters
```{r}
set.seed(123)
km.out <- kmeans(dat.scaled2, 2, nstart = 25)
km.clusters = km.out$cluster

hc.out <- hclust(dist(dat.scaled2), method = "ward.D2")
hc.clusters = cutree(hc.out, 2)

table(km.clusters, hc.clusters)

aggregate(completed_dat2, by = list(cluster = hc.clusters), mean)
```
## compare kmedoids from hierarchical
```{r}
pam.clusters <- pam.res$clustering
table(pam.clusters, hc.clusters)
```

## perform hierarchical clustering on first three PCs
```{r}
hc.out <- hclust(dist(pr.out$x[, 1:3]), method = "ward.D2")
plot(hc.out)


# beautiful neuron
library(igraph)
fviz_dend(hc.out, k = 5, k_colors = "jco",
          type = "phylogenic", repel = TRUE)
```
# Divisive Hierarchical Clustering 
```{r}

# compute it with diana function; no need to specify the method
hc4 <- diana(dat.scaled2)
hc4$dc ##0.9001963
fviz_dend(hc4, k = 2, k_colors = "jco",
          type = "phylogenic", repel = TRUE)
# plot the dendrogram
pltree(hc4, cex = 0.6, hang = -1, main = "Dendrogram of Diana", xlab = "", sub = "")
fviz_dend(hc4, k = 2, k_colors = "jco",
          type = "phylogenic", repel = TRUE)
# cut diana() tree into 2 groups
diana.clusters <- cutree(as.hclust(hc4), k = 2)
table(km.clusters, diana.clusters)
table(pam.clusters, diana.clusters)
# using ggplot
dend <- completed_dat2 %>% 
  scale %>% 
  diana() %>% 
  as.dendrogram

dend %>% set("branches_k_color", k = 2) %>% 
  plot(main = "Dendrogram of Diana", leaflab = "none")
abline(h = 8, lty = 3)
```
## compare hierarchical clustering with complete vs ward method
```{r}
# Compute distance matrix
res.dist <- dist(dat.scaled2, method = "euclidean")

# Compute 2 hierarchical clusterings
hc1 <- hclust(res.dist, method = "complete")
hc2 <- hclust(res.dist, method = "ward.D2")

# Create two dendrograms
dend1 <- as.dendrogram (hc1)
dend2 <- as.dendrogram (hc2)

tanglegram(dend1, dend2)

# customise tanglegram
dend_list <- dendlist(dend1, dend2)

tanglegram(dend1, dend2,
           highlight_distinct_edges = FALSE, # Turn-off dashed lines
           common_subtrees_color_lines = TRUE, # Turn-off line colors
           common_subtrees_color_branches = TRUE, # Color common branches 
           highlight_branches_lwd = FALSE,
           main = paste("Entanglement =", round(entanglement(dend_list), 2)),
           lwd = 0.8,
           main_left = "Complete Method",
           main_right = "Ward Method",
           lab.cex = NULL,
) %>% set("by_labels_branches_lwd", 0.5)


# compare 2 clusters obtained from hierarchical clustering and K-means clustering

set.seed(2)
km.out <- kmeans(dat.scaled2, 2, nstart = 20)
km.clusters <- km.out$cluster
table(km.clusters, hc.clusters)

# correlation-based distance
dd <- as.dist(1-cor(t(xsc)))


plot(hclust(dd, method = "complete"), main = "Complete Linkage with Correlation-Based Distance", xlab = "", sub = "", cex = .9)

##6.2 on first 3 PCs ----
hc.out=hclust(dist(pr.out$x[,1:3]))

plot(hc.out,  main="Hierarchical Clustering on First
Three Score Vectors ", sub = "", xlab = "")

abline(h = 5, col = "red", lty = 3,lwd = 3)
```

# BIC based modelling
```{r}
library(mclust)
mc1 <- Mclust(dat.scaled2, G = 1)
mc1$bic = -5635.771

mc2 <- Mclust(dat.scaled2, G = 2)
mc2$bic #-5651.584 # (EEE,2) 
summary(mc2)

mc3 <- Mclust(dat.scaled2, G = 3)
mc3$bic # -5666.159 # (EEE,3) 
summary (mc3)

mc4 <- Mclust(dat.scaled2, G = 4)
mc4$bic # -5696.416 # (EEE,4) 

mc5 <- Mclust(dat.scaled2, G = 5)
mc5$bic # -5726.493 # (EEE,5) 


mc6 <- Mclust(dat.scaled2, G = 6)
mc6$bic # -5755.349

mc7 <- Mclust(dat.scaled2, G = 7)
mc7$bic # -5775.658
summary(mc7)

mc8 <- Mclust(dat.scaled2, G = 8)
mc8$bic # -5800.048 
summary(mc8)

mc9 <- Mclust(dat.scaled2, G = 9)
mc9$bic # -5800.048 
summary(mc9)
# G = 2 has the highest BIC value with the model option EEE, followed by 3, 4, 5

BIC <- mclustBIC(dat.scaled2)
plot(BIC)
summary(BIC)
# Best BIC values:
#EEE,1     EEV,1     EVE,1
#BIC      -5635.771 -5635.771 -5635.771
#BIC diff     0.000     0.000     0.000

mod1 <- Mclust(dat.scaled2, x = BIC)
summary(mod1, parameters = TRUE)
# log-likelihood   n df       BIC       ICL
#-2735.714 440 27 -5635.771 -5635.771

ICL <- mclustICL(dat.scaled2)
summary(ICL)


# Best ICL values:
#EEE,1     EEV,1     EVE,1
#ICL      -5635.771 -5635.771 -5635.771
#ICL diff     0.000     0.000     0.000
LRT <- mclustBootstrapLRT(dat.scaled2, modelName = "EEE")
LRT
mc2$classification
plot(mc2, what = "classification")
plot(mc2, what = "density")
table(mc2$classification)
# EEE means that the clusters have the same volume, shape and orientation in p-dimensional space

library(mclust)


G <- c(2,3,4,5,6)
modelNames <- c("EII", "VII", "EEI", "VEI", "EVI", "VVI", "EEE", "EEV", "VEV", "VVV")

nnr <- length(G)*length(modelNames)
resu  <- array(data=NA, dim=c(nnr,3), dimnames=list(paste(1:nnr),c("G", "modelNames", "BIC")))

counter <- 1
for (i in G){
  for(j in modelNames){
    mc_option <- Mclust(dat.scaled2, G = i, modelNames = j)
    resu[counter, 1] <- as.numeric(i)
    resu[counter, 2] <- paste(j)
    resu[counter, 3] <- as.numeric(mc_option$BIC)
    if (counter < nrow(resu)) counter <- counter+1
  }
}

G <- as.numeric(resu[,1])
bic <- as.numeric(resu[,3])
model <- resu[,2]
dat <- data.frame(G,bic, model)

aa <- subset(dat, bic == max(bic))
aa #  G       bic model
#7 2 -5651.584   EEE

mc <- Mclust(dat.scaled2, G = 2, modelNames = "EEE") # Model-based-clustering
summary(mc)
table(hc.clusters, mc$classification)
plot(mc, what = "classification")
plot(mc, what = "uncertainty")


fviz_cluster(mc, data = dat.scaled2,
             palette = c("#FFBF00", "#29BF12"),
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = TRUE, # Avoid label overplotting (slow)
             ggtheme = theme_minimal() 
) +labs(x = "First Principal Component (48.7%)", y =  "Second Principal Component (17.4%)") + ggtitle("Model-based Clustering") + theme(plot.title = element_text(h = 0.5, size = 12, face = "bold")) 

fviz_cluster(mc, data = dat.scaled2, axes = c(2, 3),
             palette = c("#FFBF00", "#29BF12"),
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = TRUE, # Avoid label overplotting (slow)
             ggtheme = theme_minimal() 
) + labs(x = "Second Principal Component (17.4%)", y =  "Third Principal Component (16.0%)") + ggtitle("Model-based Clustering") + theme(plot.title = element_text(h = 0.5, size = 12, face = "bold")) 
```
# optimal model
```{r}
mc_ooptimal <- Mclust(dat.scaled2)
summary(mc_ooptimal)
plot(mc_ooptimal, what = "BIC")
plot(data = dat, x = as.factor(G), y = bic, type = "l")
```

# Cluster Validation 
```{r}
library(clValid)
clmethods <- c("hierarchical","kmeans","pam", "diana", "model")
intern <- clValid(dat.scaled2, nClust = 2:6,
                  clMethods = clmethods, validation = "internal")
# Summary
summary(intern)

layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))
plot(intern)

# Stability measures
clmethods <- c("hierarchical","kmeans","pam", "diana", "agnes", "model")
stab <- clValid(dat.scaled2, nClust = 2:6, clMethods = clmethods,
                validation = "stability")
               
```